Is the direction of Granger causal influence the same as the direction of information flow?

### 2015-07-07 paper.tex
Add conclusions and discussions section.

\begin{enumerate}
	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation. Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}. A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception. While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}. Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric. However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials. {\color{red} (Move to conclusions)}
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
	\item We have also restricted ourselves to analyzing linear feedback communication strategies. In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}. In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear. This could be scope for future work. However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
\end{enumerate}

@@

There are several shortcomings of this work that need to be addressed for a more comprehensive understanding of the issue. First, our information source does not evolve with time, and hence our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source. Second, our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work. Third, the power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension. Finally, we consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.

We note that Granger's original analysis does not compare Granger causality on forward and reverse links. We are doing so, in part because it is often done in practice. It strikes us that this is being done because of the ``conservation of directed information'', an equality first derived by Massey~\cite{massey2005conservation}. This equality, which states that the sum of forward and reverse Directed Informations is the mutual information, could be viewed as suggesting that an increase in Directed Information in one direction must reduce it in the other direction. However, this is not true. For e.g., making $X$ and $Y$ uncorrelated reduces Directed Information in both directions (along with their sum) to zero. In other words, Massey's law is not a conservation law in the sense of conservation laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities.

>>>>

\begin{enumerate}
	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation. Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}. A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception. While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}. Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
	\item We have also restricted ourselves to analyzing linear feedback communication strategies. In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}. In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear. This could be scope for future work. However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
\end{enumerate}

@@

We {\color{red} show(?)} that causal influence and information flow are distinct quantities. We demonstrate, by means of a concrete counter-example, that the direction predicted by causal influence metrics such as Granger Causality and Directed Information can be opposite to the true direction of information flow. There are, however, several shortcomings to our analysis, which we list in section~\ref{sec:objections-and-shortcomings}. We seek to address many of these shortcomings in future work.

It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the system parameters of the model directly to compute the Granger Causality metric. However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.

As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}. We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making {\color{red} (what might be construed as hopeful)} extrapolations from causal influences as to the manner in which information flows.

We do not seek to understate the importance of determining causal influences in the brain; understanding causal influence itself may have a great deal of benefit. For instance, we might seek to understand the spread of activity in the brain during an epileptic siezure -- here, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.

%We note that Granger's original analysis does not compare Granger causality on forward and reverse links. We are doing so, in part because it is often done in practice. It strikes us that this is being done because of the ``conservation of directed information'', an equality first derived by Massey~\cite{massey2005conservation}. This equality, which states that the sum of forward and reverse Directed Informations is the mutual information, could be viewed as suggesting that an increase in Directed Information in one direction must reduce it in the other direction. However, this is not true. For e.g., making $X$ and $Y$ uncorrelated reduces Directed Information in both directions (along with their sum) to zero. In other words, Massey's law is not a conservation law in the sense of conservation laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities.
